New modal operators for constructive mathematics

### 2023-05-02 slides-luminy2023.tex
Final touches

\addtocounter{framenumber}{-1}

@@

  \begin{tikzpicture}[remember picture,overlay]
    \node[thick, fill=black, rectangle, inner sep=0.3em, right=2em of onlyclass] (moral) {
      \begin{minipage}{6cm}
        \begin{columns}
          \begin{column}{0.15\textwidth}
            \hspace*{1.2em}\color{white}\dbend
          \end{column}
          \begin{column}{0.95\textwidth}
            \color{white}\footnotesize
            \it Don't quantify over points of spaces which might not have enough.
          \end{column}
        \end{columns}
      \end{minipage}
    };
    \path[draw=red!90,thick,-stealth]
      (moral) to
      [out=180, in=-00] (onlyclass.east);
  \end{tikzpicture}
  \pause

@@

  \emph{Can we extract a constructive proof that well implies well-founded?}
\end{frame}

@@

    \item \good{Yes}, if~$A$ is countable (irrespective of membership decidability):
    {\footnotesize\begin{align*}
      \mmm_0 &= \{ 0 \}, &
      \mmm_{n+1} &= \mmm_n + (\underbrace{\{ x \in A \,|\, x = x_n \wedge
        1 \not\in \mmm_n + (x_n) \}}_{\text{a certain subsingleton set}})
    \end{align*}}
    \vspace*{4.5em}
    \visible<11>{{
      \centering
      \raisebox{0pt}[0pt][0pt]{
        \hspace*{1.5em}
        \begin{tikzpicture}
          \node (inner) at (17.3mm,-20mm) {\textit{``a bad joke''}};
          \path (0,0) pic{laurel-wreath};
        \end{tikzpicture}
        \hspace*{-3em}
        \begin{tikzpicture}
          \node (inner) at (17.5mm,-18mm) {\vbox{\small\centering\textit{``non- \\informative''}}};
          \path (0,0) pic{laurel-wreath};
        \end{tikzpicture}
      }
      \par
    }}
    \pause
    \pause
    \vspace*{-4.5em}

@@

\begin{frame}{Questions}
  \begin{enumerate}
    \item \!Why was the inductive revolution so powerful?
    \bigskip

@@

  \hil{$\blacktriangleright$}
  For \hil{positive open} extensions, they are equivalent for bounded first-order formulas. \\
  \phantom{\hil{$\blacktriangleright$}} {\footnotesize -- e.g. $V^\nabla$ containing
  the generic surjection~$\NN \twoheadrightarrow X$, if~$X$ is inhabited.\par}
  \bigskip}
  \pause

  \textbf{Def.} A statement~$\varphi$ holds \ldots
  \begin{enumerate}
    \small
    \item[$\blacktriangleright$] \hil{everywhere} ($\necessary\varphi$) iff
    it holds in every extension.
    \item[$\blacktriangleright$] \hil{somewhere} ($\possible\varphi$) iff
    it holds in some positive extension.
    \item[$\blacktriangleright$] \hil{proximally} ($\xpossible\varphi$) iff
    it holds in some positive overt extension.
  \end{enumerate}
  \bigskip
  \pause

  \small
  \begin{multicols}{2}
    For every set~$X$, \emph{proximally} \\
    there is an enumeration~$\NN \twoheadrightarrow X$.

    A preorder is well iff \emph{everywhere}, \\
    every sequence is good.

    A ring element is nilpotent iff \\
    all prime ideals \emph{everywhere} contain it.

    For every ring, \emph{proximally} \\ there is a maximal ideal.

    A relation is well-founded iff \emph{everywhere}, there is no descending chain.

@@

\begin{frame}{The multiverse perspective}
  \begin{enumerate}
    \item \!\emph{Why was the inductive revolution so powerful?}

@@

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\includegraphics[width=\paperwidth]{eventually-monad-cropped}\end{minipage}}
\begin{frame}
  \vspace*{\paperheight}\vspace*{-2.05em}\centering
  \emph{Agda formalization available.}
\end{frame}}

>>>>

\addtocounter{framenumber}{-2}

{
\definecolor{mypurple}{RGB}{150,0,255}
\setbeamercolor{structure}{fg=mypurple}
\begin{frame}{}
  \begin{center}\includegraphics[width=0.2\textwidth]{eigenvector}\end{center}

  Let a continuous family of symmetric matrices be given:
  \[
  \begin{pmatrix}a_{11}(t)&\cdots&a_{1n}(t)\\\vdots&&\vdots\\a_{n1}(t)&\cdots&a_{nn}(t)\end{pmatrix}
  \]

  There for every parameter value~$t$, classically there is

  \hil{$\blacktriangleright$} a full list of eigenvalues~$\lambda_1(t),\ldots,\lambda_n(t)$ and \\
  \hil{$\blacktriangleright$} an eigenvector basis~$(v_1(t),\ldots,v_n(t))$.
  \bigskip

  \begin{columns}[c]
    \begin{column}{0.01\textwidth}
      \includegraphics[height=2.4em]{question-mark}
    \end{column}
    \begin{column}{0.9\textwidth}
      \mbox{Can locally the functions~$\lambda_i$ be chosen to be continuous?
      \only<2->{\hil{Yes.}}} \\
      How about the~$v_i$? \only<2->{\hil{No.}}
    \end{column}
  \end{columns}
\end{frame}
}

@@

  \begin{tikzpicture}[remember picture,overlay]
    \node[thick, fill=black, rectangle, inner sep=0.3em, right=2em of onlyclass] (moral) {
      \begin{minipage}{6cm}
        \begin{columns}
          \begin{column}{0.15\textwidth}
            \hspace*{1.0em}\color{white}\dbend
          \end{column}
          \begin{column}{0.95\textwidth}
            \color{white}\footnotesize
            \it Don't quantify over points of spaces which might not have enough.
          \end{column}
        \end{columns}
      \end{minipage}
    };
    \path[draw=red!90,thick,-stealth]
      (moral) to
      [out=180, in=-00] (onlyclass.east);
  \end{tikzpicture}
  \pause

@@

  Can we extract a constructive proof that well preorders are well-founded?
\end{frame}

@@

    \item \good{Yes}, if~$A$ is countable (irrespective of membership decidability):
    {\footnotesize\begin{align*}
      \mmm_0 &= \{ 0 \}, &
      \mmm_{n+1} &= \mmm_n + (\underbrace{\{ x \in A \,|\, x = x_n \wedge
        1 \not\in \mmm_n + (x_n) \}}_{\text{a certain subsingleton set}})
    \end{align*}}
    \vspace*{3.5em}
    \visible<11>{{
      \centering
      \raisebox{0pt}[0pt][0pt]{
        \hspace*{1.5em}
        \begin{tikzpicture}
          \node (inner) at (17.3mm,-20mm) {\textit{``a bad joke''}};
          \path (0,0) pic{laurel-wreath};
        \end{tikzpicture}
        \hspace*{-3em}
        \begin{tikzpicture}
          \node (inner) at (17.5mm,-18mm) {\vbox{\small\centering\textit{``non- \\informative''}}};
          \path (0,0) pic{laurel-wreath};
        \end{tikzpicture}
      }
      \par
    }}
    \pause
    \pause
    \vspace*{-3.5em}

@@

\begin{frame}{Questions}
  \begin{enumerate}
    \item \!Why has the inductive revolution been so powerful?
    \bigskip

@@

  \hil{$\blacktriangleright$}
  For \hil{positive overt} extensions, they are equivalent for bounded first-order formulas. \\
  \phantom{\hil{$\blacktriangleright$}} {\footnotesize -- e.g. $V^\nabla$ containing
  the generic surjection~$\NN \twoheadrightarrow X$, if~$X$ is inhabited.\par}
  \bigskip}
  \pause

  \textbf{Def.} A statement~$\varphi$ holds \ldots
  \begin{enumerate}
    \small
    \item[$\blacktriangleright$] \hil{everywhere} ($\necessary\varphi$) iff
    it holds in every extension.
    \item[$\blacktriangleright$] \hil{somewhere} ($\possible\varphi$) iff
    it holds in some positive extension.
    \item[$\blacktriangleright$] \hil{proximally} ($\xpossible\varphi$) iff
    it holds in some positive overt extension.
  \end{enumerate}
  \bigskip

  \visible<2>{\begin{tikzpicture}[overlay]
    \node[anchor=south east,inner sep=0] (image) at (14.8,-0.8) {
      \includegraphics[width=0.2\textwidth]{branching}
    };
  \end{tikzpicture}\vspace*{-1.6em}}
  \visible<3>{\begin{tikzpicture}[overlay]
    \node[anchor=south east,inner sep=0] (image) at (14.8,0.9) {
      \includegraphics[width=0.2\textwidth]{multiverse}
    };
  \end{tikzpicture}\vspace*{-0.3em}}

  \only<2>{\scriptsize Foreshadowed by:
  \begin{itemize}
    \scriptsize
    \item[1984\phantom{s}] André Joyal, Miles Tierney. \emph{An extension of the Galois theory of
    Grothendieck.}
    \item[1987\phantom{s}] Andreas Blass. \emph{Well-ordering and induction in intuitionistic
    logic and topoi.}
    \item[2010s] Milly Maietti, Steve Vickers. Ongoing work on arithmetic
    universes.
    \item[2011\phantom{s}] Joel David Hamkins. \emph{The set-theoretic multiverse.}
    \item[2013\phantom{s}] Shawn Henry. \emph{Classifying topoi and preservation of higher
    order logic by geometric morphisms.}
  \end{itemize}}
  \pause

  \small
  \begin{multicols}{2}
    For every set~$X$, \emph{proximally} \\
    there is an enumeration~$\NN \twoheadrightarrow X$.
    \bigskip

    A preorder is well iff \emph{everywhere}, \\
    every sequence is good.
    \bigskip

    A ring element is nilpotent iff \\
    all prime ideals \emph{everywhere} contain it.
    \bigskip

    For every ring, \emph{proximally} \\ there is a maximal ideal.
    \bigskip

    A relation is well-founded iff \emph{everywhere}, there is no descending chain.
    \bigskip

@@

\begin{frame}{The multiverse perspective}
  \begin{enumerate}
    \item \!\emph{Why has the inductive revolution been so powerful?}

@@

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\includegraphics[width=\paperwidth]{eventually-monad-cropped}\end{minipage}}
\begin{frame}
  \vspace*{\paperheight}\vspace*{-2.05em}\centering
  \emph{Partial Agda formalization available.}
\end{frame}}
